Fail Faster

### 2025-03-15 paper/paper.tex
lil cut

\jwc{``Property based testing is a race against time...''}
\\
\tr{Property-based testing (PBT) uses generators of random data to produce inputs to a
system under test, which is then checked against a set of properties the system is
 expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
 to software correctness, aimed at detecting bugs within a short testing 
 window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
 it is 
 imperative that inputs be generated as fast as possible; in other words, 
 generators must be treated as performance-sensitive code.}

\tr{The locus for improvement in this area is \textit{generator libraries}: if programmers are 
 provided with the tools to write fast generators, then they do not themselves 
 need to be performance experts. However, existing libraries fail to achieve good performance 
 due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
 of selecting random values. The first is random sampling, or the process of selecting random values. 
 The second is abstraction: while generator libraries are designed for flexibility and expressiveness, 
 this comes at the cost of performance, resulting in frequent boxing and unboxing and unnecessary 
 allocations (which lead to costly garbage collection pauses). In strict functional languages, 
 constructs like monadic bind introduce function call overhead, requiring each step in a generator 
 sequence to allocate closures and thread state through function applications.
 }

@@

The RNG sits at the heart of the hot path.
Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
and generator combinators like \texttt{list} call those basic generators. \hg{To
generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
\proposecut{we argue that} existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

\hg{This paragraph feels a bit clumsy right now. What if we started with a table
of the RNGs used by a bunch of popular PBT frameworks and then argued that
things like Lehmer are faster?}
For example, significantly faster RNG algorithms than SplitMix or JSF exist,
such as the Lehmer RNG \cn. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
% https://github.com/lemire/testingRNG/tree/master
Moreover, PBT libraries could even consider eschewing the requirement that a source of
entropy pass statistical tests like BigCrush\cn.\hg{This needs more discussion}  This is common practice in other areas of testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
pointer and reading from memory (which can be pipelined trivially) will always
be faster on modern CPUs than any RNG with data-dependent arithmetic
instructions.\hg{Should we just always be doing this? I think we want to argue
that this is a bridge too far because you still need to generate the buffer
ahead of time and you can run out of randomness?}

\hg{Alternative framing: ``Of course, simply arguing that the RNG is on the hot
path does not guarantee that a faster RNG leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the RNG for \bq\ with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}
In the context of this paper, however, clearly demonstrating the benefits of any of the
above faster options is tricky. Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally\hg{Is this how we're explaining this? I'm not sure that's quite right} identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property. In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.

>>>>

\jwc{``Property based testing is a race against time...''}

@@

The RNG sits at the heart of the hot path.
Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
and generator combinators like \texttt{list} call those basic generators.
\hg{To
generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
existing PBT libraries make relatively inefficient choices on this front,
leading to worse bugfinding power than what is possible.

\hg{This paragraph feels a bit clumsy right now. What if we started with a table
of the RNGs used by a bunch of popular PBT frameworks and then argued that
things like Lehmer are faster?}

For example, significantly faster RNG algorithms than SplitMix or JSF exist,
such as the Lehmer RNG \cn. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
% https://github.com/lemire/testingRNG/tree/master
Moreover, PBT libraries could even consider eschewing the requirement that a source of
entropy pass statistical tests like BigCrush\cn.\hg{This needs more discussion}  This is common practice in other areas of testing already:
fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
pointer and reading from memory (which can be pipelined trivially) will always
be faster on modern CPUs than any RNG with data-dependent arithmetic
instructions.\hg{Should we just always be doing this? I think we want to argue
that this is a bridge too far because you still need to generate the buffer
ahead of time and you can run out of randomness?}

\hg{Alternative framing: ``Of course, simply arguing that the RNG is on the hot
path does not guarantee that a faster RNG leads to measurably faster
generation; for that, we need an experiment. The most obvious experiment is to
simply swap out the RNG for \bq\ with a totally different, faster one. But, as
discussed previously, generators with different generation orders can be
difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}

In the context of this paper, however, clearly demonstrating the benefits of any of the
above faster options is tricky. Our bug-finding evaluation's power is predicated on
the fact that all of the different versions of a generator we test are
extensionally\hg{Is this how we're explaining this? I'm not sure that's quite right} identical: for a given seed and size parameter, they all produce
the \emph{same} value.  Using a different random number generator or different
source of entropy would break this property. In Section~\ref{eval}, we
demonstrate that faster RNG matters for bugfinding by exploting a ``natural
experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
can be improved \emph{without} changing its extensional behavior.
In particular,
due to implementation details related to the OCaml garbage collector,
values of the OCaml type \texttt{int64} are not machine words, but rather
\emph{pointers} to machine words. This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output, which has a significant performance benefit.
By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
performance can be improved just by using a more performant RNG.
